Artificial inteligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2023-09-29 report/markdown/report.Rmd
Figures, Tables and complete Analysis Report

@@

@@

@@

@@

```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Frequencies of any chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulations and consolidations at the consecutive follow-ups after COVID-19 in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
</p>

@@

```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Severity of chest computed tomography (CT) abnormalities was assessed by a human-determined CT severity score (CTSS), and AI-determined percentages of the lung tissue with opacity and high opacity. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
</p>

@@

## Lung function testing paramaters and abnormalities

@@

```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Frequency of any lung function testing (LFT) abnormalities, as well as percentage of patients with DLCO, FVC and FEV1 values below 80% of the reference values at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second. 
</p>

@@

```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-numeric-course). Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Lung function was assessed by measurement of DLCO, FVC and FEV1 expressed as percentage of the age-, weight- and sex-specific reference. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.
</p>

@@

@@

# Detection of radiological lung findings with AI-determined opacity and high opacity

@@

@@

>>>>

* The current analysis of data of COVID-19 convalescents pursued three goals: (1) to investigate concordance between prediction of lung computed tomography (CT) abnormalities detected by a human radiologist and by an artificial intelligence (AI) powered software, (2) to establish a relationship between AI-determined lung opacity or high opacity and CT pathology severity assessed by the radiologist, and (3) to explore, how lung CT results, demographic and clinical features translate into impaired lung function in COVID-19 patients.

* These goals were addressed by a re-analysis of the published observation [CovILD data set](https://clinicaltrials.gov/study/NCT04416100) consisting of baseline demographic and clinical information, data on the course and treatment of acute COVID-19 as well as longitudinal assessment of persistent symptoms, lung CT scans and lung function testing (LFT) at two, three, six and twelve months after COVID-19 diagnosis [@Sonnweber2022; @Luger2022; @Sonnweber2020; @Sahanic2023].

* A total of `r nrow(lft_globals$analysis_tbl)` observations obtained from `r length(covild$complete_ids)` CovILD study participants with complete CT and LFT results were eligible for the analysis. Statistical tools used to address the analysis goals included receiver operating characteristic, bootstrapped inter-rater reliability metrics, linear, robust linear and generalized additive modeling as well as multi-parameter modeling with machine learning.

* In a comparison with the human radiologist, AI-determined lung opacity demonstrated a good accuracy at detection of any CT abnormalities, ground glass opacity (GGO) and reticulations in the lungs of COVID-19 convalescents. By contrast its ability to detect lung consolidations was poor. Performance of AI-determined high opacity at detection of any CT findings, GGO, and reticulation was moderate, its accuracy at identification of consolidation was fair.

* CT severity score (CTSS) was established as a tool to determine an overall grade of radiological lung abnormalities in COVID-19 patients by radiologists [@Luger2022; @Sonnweber2020]. As investigated by receiver-operating characteristic analysis, values of CTSS > 10 points may be used to identify cases of interstitial lung disease (> 5% of lung opacity computed by AI) with a moderate reliability. AI-computed opacity and high opacity can be associated with the radiologist's CTSS with a simple polynomial formula.

* Among major LFT readouts, diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC) and forced expiratory volume in one second (FEV1), only DLCO could be successfully modeled as a function of demographic, clinical and lung CT explanatory factors. This suggests that DLCO but not FVC or FEV1 may be impacted by SARS-CoV-2 infection-mediated lung damage.

* In a more detailed receiver operating characteristic analysis, we could show that even mild structural lung abnormalities (CTSS > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'CTSS')$cutoff[1] %>% signif(2)`, opacity > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'opacity_percent')$cutoff[1] %>% signif(2)`%, high opacity > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'high_opacity_percent')$cutoff[1] %>% signif(2)`) may translate to insufficient DLCO defined as values below 80% of the age-, sex- and weight-specific reference.

* Small size of the cohort and unavailability of a validation data set were the main limitations of the analysis. In particular, external validation of the LFT modeling and the CT severity - LFT relationship would allow for an unbiased evaluation of the model predictions and reliability of CTSS, opacity and high opacity as a markers of functional lung deficiency.

@@

# Analysis goals

The analysis pursued three main goals:

* comparison of artificial intelligence (AI) determined lung density and lung high density with human-determined radiological lung abnormalities in COVID-19 convalescents

* determination of a cutoff of the radiologist's computed tomography (CT) severity score (CTSS) to detect cases of post-COVID-19 interstitial lung disease defined as opacity in >5% of the lung tissue and development of a mathematical relationship between CTSS and AI-determined opacity and high opacity

* prediction of outcomes and abnormalities of lung function testing (LFT) in COVID-19 convalescents by radiological readouts combined with data on demographics, clinical history, course of acute COVID-19, time afer COVID-19 diangosis, persistent respiratory symptoms and physical performance rating

```{r fig-analysis-strategy, fig.width = figur::convert(figures$analysis_strategy, to = 'in')$w, fig.height = figur::convert(figures$analysis_strategy, to = 'in')$h, fig.cap = 'Goals and strategy of the analysis.'}

figures$analysis_strategy$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-analysis-strategy). Goals and strategy of the analysis.</b>
<br>
AI: artificial intelligence; CT: computed tomography; LFT: lung function testing; CTSS: radiologist's CT severity score; ROC: receiver-operating characteristic; ILD: interstitial lung disease; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; SHAP: Shapley additive explanations.
</p>

@@

Herein, baseline demographic and clinical characteristic, data on the course and treatment of acute COVID-19, longitudinally recorded lung CT and LFT measurements as well as longitudinally surveyed respiratory symptoms and physical performance rating in the published CovILD study were investigated [@Sahanic2023; @Luger2022; @Sonnweber2022; @Sonnweber2020]. 
In brief, the longitudinal observation CovILD study cohort consisted of convalescents of symptomatic SARS-CoV-2 infection during the fist outbreak of the pandemic. 
Initially, `r length(covild$common_ids)` individuals were enrolled between March 2023 and June 2023 from three clinical centers in Tyrol, Austria: the University Clinic of Innsbruck, St. Vinzenz Hospital in Zams and Karl Landsteiner Acute Rehabilitation Center in Münster. 
The study inclusion criteria were PCR-confirmed SARS-CoV-2 infection and presence of characteristic COVID-19 symptoms. 
Follow-up visits were scheduled at two, three, six and twelve months after COVID-19 diagnosis. 
The routine follow-up visit protocol included 
general examination by a physician, 
survey of symptoms (e.g. fever, cough, sleep problems, olfactory dysfunction), 
assessment of dyspnea with the modified Medical Research Council scale (mMRC) and 
rating of physical performance with the Eastern Cooperative Oncology Group scale (ECOG) 
along with LFT, chest CT and trans-thoracic electrocardiography. 
Information on the course, severity and treatment of acute COVID-19 was obtained at the two-month follow-up via an interview of the participant and screening of the hospital electronic records [@Sahanic2023; @Sonnweber2020]. 
CT scans were evaluated by a team of experienced radiologists according to the Fleischner society guidelines. 
CT lung opacity and high lung opacity expressed as percentages of the organ were computed with an AI - powered software (Syngo.via CT Pneumonia Analysis Software) [@Luger2022; @Sonnweber2020].

In the current report, `r nrow(lft_globals$analysis_tbl)` observations from all follow-up examinations with the complete CT and LFT results recorded in `r length(covild$complete_ids)` participants were analyzed. 
Please note, that per participant, up to four observations were available. 
Due to this participant-matching, the observations were not independent, which had consequences for the analysis procedures, e.g. for multi-parameter modeling (__Figure \@ref(fig:fig-consort)__, __Table \@ref(tab:tab-study-variables)__).

The analyzed participants were grouped according to the severity of acute COVID-19 as 
convalescents of ambulatory COVID-19 (`r get_percent(cohort$analysis_tbl, 'severity_class')['ambulatory mild']` of the cohort, home-isolated during infection, WHO ordinal scale for clinical improvement: 1 - 2), 
hospitalized moderate COVID-19 patients (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized moderate']`, without oxygen or with low-flow oxygen, WHO ordinal scale for clinical improvement: 3 - 4), 
and hospitalized severe COVID-19 patients (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized severe']`, high-flow oxygen or mechanical ventilation, WHO ordinal scale for clinical improvement: 5 - 7) (__Table \@ref(tab:tab-cohort)__). 
In the analyzed data set, `r get_percent(cohort$analysis_tbl, 'sex')['male']` of participants were male. 
The percentage of males was significantly higher in moderate and severe COVID as compared with ambulatory COVID-19 individuals (`r get_test(cohort$test, 'sex')`, moderate). 
The median age at COVID-19 diagnosis was `r signif(median(cohort$analysis_tbl$age), 2)` years, ambulatory COVID-19 patients were significantly younger than hospitalized COVID-19 survivors (`r get_test(cohort$test, 'age')`, moderate). 
Ex- or current smokers made up to `r sum(get_percent(cohort$analysis_tbl, 'smoking', percent_mark = FALSE)[2:3])`% of the analyzed cohort. 
There were no significant differences in participants with smoking history between the acute COVID-19 severity subsets. 
Yet, the pack-year number was the highest in hospitalized moderate COVID-19 convalescents (`r get_test(cohort$test, 'smoking_pack_years')`, weak). 
Based on body mass index (BMI) > 25 kg/m^2^, 
`r sum(get_percent(cohort$analysis_tbl, 'body_mass_class', percent_mark = FALSE)[2:3])`% of participants were overweight or obese; there were no significant differences in percentages of obese or overweight participants between the acute COVID-19 severity subsets. 
Endocrine or metabolic conditions (`r get_percent(cohort$analysis_tbl, 'endocrine_metabolic_comorbidity')['yes']` of analyzed participants), 
cardiovascular diseases (`r get_percent(cohort$analysis_tbl, 'cardiovascular_comorbidity')['yes']`), 
hypertension (`r get_percent(cohort$analysis_tbl, 'hypertension_comorbidity')['yes']`), 
hypercholesterolemia (`r get_percent(cohort$analysis_tbl, 'hypercholesterolemia_comorbidity')['yes']`), 
and type II diabetes (`r get_percent(cohort$analysis_tbl, 'diabetes_comorbidity')['yes']`) 
were the most common comorbidities. 
Endocrine or metabolic comorbidities (`r get_test(cohort$test, 'endocrine_metabolic_comorbidity')`), 
cardiovascular conditions (`r get_test(cohort$test, 'cardiovascular_comorbidity')`), 
hypercholesterolemia (`r get_test(cohort$test, 'hypercholesterolemia_comorbidity')`), 
and diabetes (`r get_test(cohort$test, 'diabetes_comorbidity')`) 
were more significantly common in survivors of moderate or severe COVID-19 as compared with ambulatory COVID-19 individuals. 
Effects sizes of these differences were weak or moderate (__Table \@ref(tab:tab-cohort)__).

The median length of hospital stay for analyzed participants was `r signif(median(cohort$analysis_tbl$hospital_stay_days), 2)` days and was longer for severe that moderate COVID-19 patients. 
Antibiotic anti-infective treatment was applied to `r get_percent(cohort$analysis_tbl, 'antiinfective_treatment')['yes']`% of participants, this percentage was the highest in severe COVID-19 patients 
(`r get_test(cohort$test, 'antiinfective_treatment')`, large). 
Macrolides and anti-platelet drugs were administered to 
`r get_percent(cohort$analysis_tbl, 'macrolide_treatment')['yes']` and 
`r get_percent(cohort$analysis_tbl, 'antiplatelet_treatment')['yes']` of participants, respectively. 
Their frequency was higher in hospitalized than in ambulatory patients 
(macrolides: `r get_test(cohort$test, 'macrolide_treatment')`, weak; 
anti-platelet: `r get_test(cohort$test, 'antiplatelet_treatment')`, weak). 
Systematic steroid administration was not considered as a treatment of moderate or severe COVID-19 during the first European outbreak of SARS-CoV-2 in 2020. 
Hence none of the study participants received corticosteroid therapy at the peak of acute COVID-19. 
Corticosteroids were administered to 
`r get_percent(cohort$analysis_tbl, 'steroid_treatment')['yes']`% of analyzed participants with persistent pneumonia beginning from the third week after diagnosis at the discretion of the physician. 
Such sub-acute corticosteroid treatment was the most frequent in severe disease convalescents as compared with moderate or ambulatory COVID-19 (`r get_test(cohort$test, 'steroid_treatment')`, moderate). 
The median weight loss during acute COVID-19 was `r abs(median(cohort$analysis_tbl$weight_change_kg))` kg and was significantly higher in severe as compared with ambulatory or moderate COVID-19 patients 
(`r get_test(cohort$test, 'weight_change_kg')`, large) (__Table \@ref(tab:tab-cohort)__).

```{r fig-consort, fig.width = figur::convert(figures$consort, to = 'in')$w, fig.height = figur::convert(figures$consort, to = 'in')$h, fig.cap = 'Analysis inclusion scheme.'}

figures$consort$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-consort). Analysis inclusion scheme.</b>
<br>
CT: computed tomography of the chest; LFT: lung function testing; GGO: ground glass opacity; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; ECG: echocardiography; AI: artificial intelligence.
</p>

```{r tab-study-variables, tab.cap = 'Study variables'}

flextable::flextable(tables$study_variables) %>% 
  set_widths(c(2.2, 3, 2.5, 3, 5, 1.5, 1.5)) %>% 
  merge_v(1:2) %>% 
  theme_vanilla

@@

Frequency of any radiological lung abnormalities, ground glass opacity (GGO) and reticulation diagnosed by a radiologist was substantially higher in hospitalized than in ambulatory COVID-19 individuals. 
Analogically, the pace of recovery from these lung lesions was the slowest in severe COVID-19 survivors: 
among `r filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up') %>% nrow` severe COVID-19 convalescents investigated at the twelve-month follow-up, GGO were observed in 
`r get_percent(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up'), 'GGO_finding')['yes']` and reticulation was evident in `r get_percent(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up'), 'reticulation_finding')['yes']` of cases. 
Consolidation was diagnosed most commonly in hospitalized severe COVID-19 individuals at the two-month follow-up (__Figure \@ref(fig:fig-ct-finding-course)__, __Table \@ref(tab:tab-cohort-ct)__). 
A more detailed descriptive analysis of lung lesions in the CovILD cohort is provided in our recent publications [@Sahanic2023; @Luger2022].

Severity of radiological lung abnormalities was gauged with the CT severity score (CTSS) determined by the radiologist and, in parallel, by percentages of organ opacity and high opacity computed by an AI-powered software [@Luger2022; @Sonnweber2020]. 
Independently of the severity readout, the most severe or widespread lesions were found in severe COVID-19 survivors followed by moderate COVID-19 patients. 
A substantial recovery from structural lung damage could be inferred from decreasing CTSS, opacity and high opacity values at the consecutive follow-ups. 
Still, low-grade residual lesions of unknown clinical relevance were observed in majority of severe COVID-19 survivors at one year after COVID-19 diagnosis (percentage of the lungs with opacity: 
median `r median(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up')$opacity_percent) %>% signif(2)`%, 
interquartile range: `r quantile(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up')$opacity_percent, c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' to ')`%) (__Figure \@ref(fig:fig-ct-numeric-course)__, __Table \@ref(tab:tab-cohort-ct)__). 
For a detailed longitudinal analysis of structural pulmonary recovery, please refer to our published reports [@Sahanic2023; @Luger2022].

@@

```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Frequencies of any chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulations and consolidations at the consecutive follow-ups after COVID-19 in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
</p>

@@

```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Severity of chest computed tomography (CT) abnormalities was assessed by a human-determined CT severity score (CTSS), and AI-determined percentages of the lung tissue with opacity and high opacity. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
</p>

@@

## Lung function testing parameters and abnormalities

LFT abnormalities were defined as values of diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC), forced expiratory volume in one second (FEV1) or total lung capacity (TLC) below < 80% of the age-, sex- and weight-specific reference, or FEV1 to FVC ratio (FEV1:FVC) below 70% of the reference value [@Sahanic2023; @Sonnweber2022; @Sonnweber2020]. Such LFT findings were discerned in more than one third of available observations at consecutive follow-ups in the entire analyzed data set 
(two months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '2-month follow-up'), 'LFT_findings')['yes']`, 
twelve months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '12-month follow-up'), 'LFT_findings')['yes']` of observations). 
Their percentage was the highest in severe COVID-19 survivors; in this subset, a clear recovery trend was observed 
(two months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '2-month follow-up', severity_class == 'hospitalized severe'), 'LFT_findings')['yes']`, 
twelve months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '12-month follow-up', severity_class == 'hospitalized severe'), 'LFT_findings')['yes']`). 
This suggests, that in the severe COVID-19 subset, functional lung pathology was at least partly mediated by SARS-CoV-2 infection. 
In turn, no clear recovery from LFT abnormalities was evident in ambulatory and moderate COVID-19 patients. 
This may indicate that functional lung deficits in those severity subsets can be COVID-19-independent, e.g. as a result of a pre-existing diagnosed or undiagnosed pulmonary condition or represent a psychosomatic phenomenon [@Hufner2023]. 
Similar observations were made for deficits of the major LFT readouts: DLCO, FVC and FEV1, as well as values of DLCO, FVC and FEV1 expressed as percentage of the respective references (__Figures \@ref(fig:fig-lft-finding-course)__ - __\@ref(fig:fig-lft-numeric-course)__, __Table \@ref(tab:tab-cohort-lft)__). 
A more detailed results of descriptive and longitudinal analysis of LFT measurements in the CovILD cohort can be found in our recent publications [@Sahanic2023; @Sonnweber2022].

@@

```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Frequency of any lung function testing (LFT) abnormalities, as well as percentage of patients with DLCO, FVC and FEV1 values below 80% of the reference values at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second. 
</p>

@@

```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

@@

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-numeric-course). Time course of numeric lung function testing readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Lung function was assessed by measurement of DLCO, FVC and FEV1 expressed as percentage of the age-, weight- and sex-specific reference. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.
</p>

@@

## Persistent respiratory symptoms and physical performance

In the current analysis we explored persistent symptoms of COVID-19 of potential relevance for lung function and structural lung lesions: cough, dyspnea and physical performance impairment [@Sahanic2023; @Sonnweber2020]. 
Dyspnea defined by modified Medical Research Council scale (mMRC) > 0 points affected 
`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'dyspnea_symptom')['yes']` and 
`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'dyspnea_symptom')['yes']` of participants at the two and twelve follow-up, respectively. 
The highest frequency of dyspnea was found in severe COVID-19 survivors at the two-month follow-up. 
Cough was reported by 
`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'cough_symptom')['yes']` of study participants at the earliest follow-up and 
by `r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'cough_symptom')['yes']` one year after COVID-19 diagnosis. 
There were no clear differences in frequency of cough between the acute COVID-19 severity subsets. 
Impaired physical performance defined as Eastern Cooperative Oncology Group scale (ECOG) > 0 points was reported by 
`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'impaired_performance_symptom')['yes']` of 
participants at the two-month follow-up and by 
`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'impaired_performance_symptom')['yes']` of participants one year after COVID-19. 
Analogically, impaired physical performance was the most common in the severe acute COVID-19 subset (__Figures \@ref(fig:fig-symptom-course)__ - __\@ref(fig:fig-symptom-rating-course)
__, __Table \@ref(tab:tab-cohort-symptoms)__). 
For a more detailed analysis of the symptom kinetic, please refer to our recent publications [@Sahanic2023; @Sonnweber2022; @Hufner2023].

```{r fig-symptom-course, fig.width = figur::convert(figures$symptom_course, to = 'in')$w, fig.height = figur::convert(figures$symptom_course, to = 'in')$h, fig.cap = 'Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

figures$symptom_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-symptom-course). Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Intensity of self-reported dyspnea was assessed with the modified Medical Research Council scale (mMRC) and presence of dyspnea was defined as mMRC > 0. 
Presence of cough was surveyed by a single 'yes'/'no' item. 
Impairment of physical performance was assessed with the Eastern Cooperative Oncology Group scale (ECOG, high values indicate more profound impairment) and impaired physical performance was defined as ECOG > 0. 
Presence of dyspnea, cough and impairment of physical performance at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
</p>

```{r fig-symptom-rating-course, fig.width = figur::convert(figures$symptom_rating_course, to = 'in')$w, fig.height = figur::convert(figures$symptom_rating_course, to = 'in')$h, fig.cap = 'Rating of dyspnea and physical performance loss at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}

figures$symptom_rating_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-symptom-rating-course). Rating of dyspnea and physical performance loss at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
<br>
Intensity of self-reported dyspnea and impairment of physical performance was assessed with the modified Medical Research Council (mMRC) and the Eastern Cooperative Oncology Group scales (ECOG, high values indicate more profound impairment), respectively. 
Dyspnea and physical performance impairment scores are presented in stack plots, with the score value coded by the bar color. Numbers of complete observations at the follow-up examinations are displayed above the data points.
</p>

```{r tab-cohort-symptoms, tab.cap = 'Presence and rating of symptoms of relevance for lung function at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_symptoms) %>% 
  set_widths(c(2.2, 2.5, rep(3.5, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('mMRC: modified Medical Research Council dyspnea scale; ECOG: Eastern Cooperative Oncology Group scale of physical performance impairment.'), 
           ref_symbols = 'a', 
           part = 'header') %>% 
  theme_vanilla

@@

# Detection of radiological lung findings by AI-determined opacity and high opacity

@@

The first main goal of the analysis was to compare radiological lung abnormalities detected by AI-determined percentage of the lung opacity and high opacity with diagnoses made by the human radiologist. 
In the initial step of this analysis task, we sought to find cutoffs of lung opacity and high lung opacity values which discriminate between observations with and without any CT findings, GGO, reticulation or consolidation discerned by the radiologist with the optimal sensitivity and specificity. 
Such optimal cutoffs were determined by an iterative screening of opacity and high opacity values and comparison of the respective Youden's J statistic values defined as $J = Sensitivity + Specificity - 1$. 
The opacity and high opacity cutoffs corresponding to the maximum of J were chosen for stratification of the observations [@Lopez-Raton2014] (__Table \@ref(tab:tab-ai-cutoffs)__).

As presented in __Figure \@ref(fig:fig-ai-opacity-roc)__, values of AI-determined opacity ranging between 
`r filter(cut_opacity$stats$optimal_cutoff, response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding', 'consolidation_finding'))$cutoff %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of the lungs were identified as the best cutoffs for detection of any CT abnormalities, GGO, reticulation and consolidation, respectively. 
Of note, the highest optimal detection cutoff of opacity was identified for consolidations. 
All four types of CT findings were detected with a similar overall accuracy as evident from comparable values of J and area under the receiver operating characteristic curve (AUC). 
Values of AI-determined high opacity ranging between 
`r filter(cut_high$stats$optimal_cutoff, response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding', 'consolidation_finding'))$cutoff %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of the lungs were proposed as the optimal cutoffs for detection of any CT abnormalities, GGO, reticulation and consolidation. 
The highest optimal cutoff of high opacity was found for consolidation. 
Interestingly, performance of high opacity at detection of consolidations was substantially better than performance of high opacity at detection of other CT findings as evident from higher sensitivity, J and AUC values (__Figure \@ref(fig:fig-ai-high-opacity-roc)__).

@@

Canonical, unweighted Cohen's $\kappa$ metric is used to assess concordance between two raters at prediction of a categorical outcome. 
Since $\kappa$ accounts for frequency of outcomes concordantly predicted by both raters by chance, it is considered a more robust measure of model's or cutoff's performance than accuracy, specificity or sensitivity, especially in data sets with highly frequent or rare outcomes [@Cohen1960; @McHugh2012]. 
Herein we employed $\kappa$ to assess the concordance at prediction of CT findings between the human radiologist and AI-determined opacity/high opacity stratified by the optimal detection cutoffs characterized above.

In such analysis performed for all available observations, lung opacity appeared to discriminate between observations with and without any CT findings, GGO and reticulation identified by the radiologist with good and comparable reliability 
($\kappa$: `r filter(op_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding'))$estimate %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, reliability of AI-determined opacity at detection of relatively infrequent consolidations was poor 
($\kappa$ = `r  filter(op_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response == 'consolidation_finding')$estimate %>% signif(2)`). 
While reliability of opacity at identification of any radiological lung abnormalities was similar in ambulatory, moderate and severe COVID-19 survivors, the concordance at detection of GGO and reticulation was substantially better in moderate or severe than ambulatory COVID-19 patients (__Figure \@ref(fig:fig-ai-opacity-severity)__, __Table \@ref(tab:tab-ai-cutoffs-severity)__). 
Notably, reliability of AI-determined opacity at detection of any CT findings, GGO and reticulation tended to be better one-year after COVID-19 diagnosis as compared with the former follow-up examinations (__Figure \@ref(fig:fig-ai-opacity-fup)__, __Table \@ref(tab:tab-ai-cutoffs-fup)__). 
Curiously, a training effect of the human radiologists may explain this surprising phenomenon but also the possibility to consult CT scans obtained from a particular patient at earlier follow-ups.
